Global code motion of congruent instructions

### 2017-11-06 cgo/paper.tex
reorder the sections
chi

The algorithm for \GCM{} uses several common representations of the
program that we shortly describe below:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

@@

\subsection{Liveness using DJ-graph}
DJ-graph \cite{Sreedhar1996} is a data structure that augments the dominator
tree with join-edges to keep track of data-flow in a program. We use DJ-graph
and merge-sets to compute liveness of variables as illustrated in
\cite{das2012}.  It is very efficient for computing liveness and does not
require any bitvectors to be maintained for each basic-block. The underlying
simplicity of liveness computation is due to \SSA{} form, where the the values
only flow (from def to use) either through dominator edges or the join edges
(where we insert a PHI.) The DJ-graph contains both these edges which allows for
computation of merge-sets for each basic-block, i.e., a set of all basic-blocks
where the values can flow from a particular basic-block. We have implemented the
merge-set computation based on a DJ-graph as illustrated in \cite{das2005}. A
simplified version of the merge-set that we implemented is presented here:

\begin{verbatim}
// Return true if the merge-set of source node of a
// visited J-edge (incoming edge of lnode) is not
// the subset of the merge-set of lnode.
bool
stillInconsistent(Node lnode, JEdges JE,
                   Visited V, MergeSet MergeSet)
  for e in (all incoming edges to lnode):
    if e is in JE and e in V:
      if MergeSet(Source Node of e)
         is not a subset of MergeSet(lnode):
        return true
  return false
\end{verbatim}

Each basic block which has a join-edge (J-Edge) to its successor will have a
merge-set that is a subset of the merge-set of its successor \cite{das2005}. The
function \emph{stillInconsistent} returns true if this invariant is not
satisfied which allows the function \emph{constructMergeSet} to repeat.

\begin{verbatim}
// Compute a merge-set breadth first order.
bool constructMergeSet_1(BFSList B, JEdges JE,
                         DomLevel DL)
  repeat = false
  V = empty list of visited edges
  for n in B:
    for e in (all incoming edges to n):
      if e is in JE and e not in V:
        V(e) = true
        let tmp = Source Node of e
        let tnode = Target Node of e
        let lnode = NULL
        while (DL(tmp) >= DL(tnode)):
          MergeSet(tmp) = MergeSet(tmp)
                          U MergeSet(tnode)
                          U {tnode}
          lnode = tmp
          tmp = dom-parent(tmp)
        repeat = still_inconsistent(lnode, JE, MergeSet)
  return repeat
\end{verbatim}

\begin{verbatim}
// Construct merge-set of each node in
// Control-Flow-Graph G of a function.
void constructMergeSet(CFG G)
  B = Breadth First Order of G
  JE = JEdges of G
  DL = Path lengths (from root) of each node in G
  do: // Call until a fixed point is reached.
    Repeat = constructMergeSet_1(B, JE, DL)
  while (Repeat)
\end{verbatim}

We compute merge sets, of the control flow graph, which remains the same
throughout the \GCM{} transformation. For \GCM{} we only want to know if a use
operand is a kill (to compute changes in register pressure). For that we only
need to know whether the use is also required later in the execution path. A
variable is live out of a basic-block B if it is used in the merge set of B. We
compute the live-out relation on-demand when profitability of hoist/sink
candidates is to be evaluated. A simplified version of
\emph{isLiveOutUsingMergeSet} is presented here:

\begin{verbatim}
// Return true if variable A is liveout from N
bool isLiveOutUsingMergeSet(Node N, Variable A)
  if (A is defined in N)
    if (A is used outside any basic-block other than N)
      return true
    return false

  MergeSet(N) = null
  // The merge-set of N is the union of
  // merge-sets of its successors.
  for W in successors(N)
    MergeSet(N) = MergeSet(N) U MergeSet(W)

  // Iterate over all the uses of A and see
  // if any intersect with the merge-set of N.
  for T in users(A)
    B = basic_block(T)
    while (B != null and B != def_bb(A))
      if B in MergeSet(N) return true
      B = dom-parent(B)
  return false
\end{verbatim}

If a variable is not live-out of a basic-block, it still may be used later in
the basic-block. To establish whether a use of a variable is a kill we iterate
on all the subsequent instructions in a basic-block checking for uses, the
\emph{isKill} function uses \emph{isLiveOutUsingMergeSet} to find out if an
operand is killed in an instruction.

\begin{verbatim}
// Return true if op is the last use of the operand in I.
bool isKill(Operand op, Instruction I)
  if (isLiveOutUsingMergeSet(op))
    return false
  for each I1 after I in basic block of I
    if (I1 uses op)
      return false // Not a kill
  return true
\end{verbatim}

The original algorithm presented in \cite{das2012} has mismatched types in terms
of uses and nodes because each node can have many instructions and hence many
uses. Also, while iterating on the dominator of each user of a variable we may
reach to the beginning of a function, in that case the inner while loop needs to
terminate. These two cases were missing from the original algorithm.

@@

@@

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
After the legality checks have passed, we check if a \gcm{} is profitable.  That
takes into account the impact \gcm{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance
across popular benchmarks and at the same time to reduce code size as much as
possible. Following cost models are implemented:

>>>>

The algorithm for \GCM{} uses several common representations of the
program that we shortly describe below:
\begin{itemize}[leftmargin=*,topsep=0pt]
\item Dominance (DOM) and Post-Dominance (PDOM) relations \cite{dragonbook} on a
  Control Flow Graph (\CFG{});
\item DJ-Graph \cite{Sreedhar1996} is a data structure that augments the
  dominator tree with join-edges to keep track of data-flow in a program. We use
  DJ-Graph to compute liveness of variables as illustrated in \cite{das2012};
\item Static Single Assignment (\SSA{}) \cite{cytron};
\item Global Value Numbering (\GVN{}) \cite{rosen1988global,click1995global}: to
  identify similar computations compilers use \GVN{}.  Each expression is given
  a unique number and the expressions that the compiler can prove to be
  identical are given the same number;
\item Rank of an expression \cite{rosen1988global}: The expressions are ordered
  according to the amount of dependency they have on other expressions.  For
  example in the expression $a = b+c$, $a$ will have higher rank than rank of
  $b$, and $c$.
\item \MemorySSA{} \cite{novillo2007memory}: it is a factored use-def chain of
  memory operations that the compiler is unable to prove
  independent. \MemorySSA{} accelerates the access to the alias analysis
  information.
\end{itemize}

@@

@@

\section{Code hoisting of congruent computations}
Examples to show CHI insertion.

The algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the ANTIC
points in the CFG which do not change during hoisting. The ANTIC points are
basically the dominance-frontiers in the inverse graph. So we introduce a data
structure (CHI nodes) to keep track of values flowing out of a basic block. We
only do this for values with multiple occurrences in the function as they are
the potentially hoistable candidates. This approach allows us to hoist
instructions to a basic block with more than two successors, as well as deal
with infinite loops in a trivial way.

Limitations: This pass does not hoist fully redundant expressions because they
are already handled by GVN-PRE. It is advisable to run gvn-hoist before and
after gvn-pre because gvn-pre creates opportunities for more instructions to be
hoisted.

\begin{figure}[h!]
  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}}
\label{fig:chi-intro}
\end{figure}

A CHI keeps information about values flowing out of a basic block.  It is
similar to PHI but in the inverse graph, and used for outgoing values on each
edge. For conciseness, it is computed only for instructions with multiple
occurrences in the CFG because they are the only hoistable candidates.

The global value number for both I1 and I2 is V, the CHI node will save the
instruction as well as the edge where the value is flowing to.  Where CHIArg1,
CHIArg2 are of type CHIArg.

\begin{verbatim}
struct CHIArg {
  // Edge destination (shows the direction of flow), may not be where the I is.
  BasicBlock *Dest;
  // The instruction which uses the values flowing out of CHI.
  Instruction *I;
  // The global value number of I.
  GVNType VN;
};

\end{verbatim}

Essentially, CHIArg1 = {V, B, I1} and CHIArg2 = {V, C, I2}

For sinking we use the similar approach as \ref{SSAPRE} to track values flowing
into a basic block.

\subsection{Profitability check (Cost models)}
\label{subsec:cost-models}
TODO: Motivate why this is important.
After the legality checks have passed, we check if a \gcm{} is profitable.  That
takes into account the impact \gcm{} would have on various parameters that might
affect runtime performance, e.g., impact on live-range, gain in the code
size. The current implementation makes effort to not regress in performance
across popular benchmarks and at the same time to reduce code size as much as
possible. Following cost models are implemented:

@@
